---
title: "Repeated Measurements"
author: "Johan Aparicio"
output: rmarkdown::html_vignette
code_folding: hide
vignette: >
  %\VignetteIndexEntry{Repeated Measurements}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `extract_rcov()` function is a practical tool for extracting the residual variance-covariance matrix from a repeated measurement ASReml model. This function is particularly useful when dealing with various covariance structures, including but not limited to the uniform correlation (corv), power or exponential (expv), antedependence (ante) and unstructured (US).

Currently, the structures available are: 

1. Simple correlation model (`corv`); homogeneous variance form.
2. Simple correlation model (`corh`); heterogeneous variance form.
3. General correlation model (`corgh`); heterogeneous variance form.
4. Exponential (or power) model (`expv`); homogeneous variance form.
5. Exponential (or power) model (`exph`); heterogeneous variance form.
6. Autoregressive model of order 1 (`ar1v`); homogeneous variance form.
7. Autoregressive model of order 1 (`ar1h`); heterogeneous variance form.
8. Antedependence variance model of order 1 (`ante`).
9. Unstructured variance model (`us`).

[Watch the tutorial](https://www.youtube.com/watch?v=BwvTFZDejXI): A good guide on fitting repeated measurement models in ASReml by VSNi. However, it might leave you wondering how to actually extract the fitted residual variance-covariance matrix. That's where `extract_rcov()` comes into play.

This vignette utilizes the same dataset featured in the video and incorporates a segment of the code to showcase the functionality of `extract_rcov()`. Additionally, we provide insightful figures that aid in exploring the results.

> To run this vignette, ensure you have an ASReml license.

```{r, eval=FALSE}
library(ggpubr)
library(agriutilities)
library(tidyr)
library(dplyr)
library(tibble)
library(asreml)

head(grassUV) |> print()
grassUV |>
  ggplot(
    aes(x = Time, y = y, group = Plant, color = Plant)
  ) +
  geom_point() +
  geom_line() +
  facet_wrap(~Tmt) +
  theme_minimal(base_size = 15)
```

```{r, warning=FALSE, message=FALSE, fig.width = 9, echo=FALSE}
library(ggpubr)
library(agriutilities)
library(data.table)
library(tidyr)
library(dplyr)
library(tibble)

if (requireNamespace("asreml", quietly = TRUE)) {
  library(asreml)
  head(grassUV) |> print()
  grassUV |>
    ggplot(
      aes(x = Time, y = y, group = Plant, color = Plant)
    ) +
    geom_point() +
    geom_line() +
    facet_wrap(~Tmt) +
    theme_minimal(base_size = 15)
}
```

## Exploration
```{r, eval = FALSE}
tmp <- grassUV |>
  group_by(Time, Plant) |>
  summarise(mean = mean(y, na.rm = TRUE)) |>
  spread(Time, mean) |>
  column_to_rownames("Plant")

a <- covcor_heat(matrix = cor(tmp), legend = "none", size = 4.5) +
  ggtitle(label = "Pearson Correlation")

b <- tmp |>
  cor(use = "pairwise.complete.obs") |>
  as.data.frame() |>
  rownames_to_column(var = "Time") |>
  gather("Time2", "corr", -1) |>
  type.convert(as.is = FALSE) |>
  mutate(corr = ifelse(Time < Time2, NA, corr)) |>
  mutate(Time2 = as.factor(Time2)) |>
  ggplot(
    aes(x = Time, y = corr, group = Time2, color = Time2)
  ) +
  geom_point() +
  geom_line() +
  theme_minimal(base_size = 15) +
  color_palette(palette = "jco") +
  labs(color = "Time", y = "Pearson Correlation") +
  theme(legend.position = "top")

ggarrange(a, b)
```

```{r, warning=FALSE, message=FALSE, fig.width = 9, echo = FALSE}
if (requireNamespace("asreml", quietly = TRUE)) {
  tmp <- grassUV |>
    group_by(Time, Plant) |>
    summarise(mean = mean(y, na.rm = TRUE)) |>
    spread(Time, mean) |>
    column_to_rownames("Plant")

  a <- covcor_heat(matrix = cor(tmp), legend = "none", size = 4.5) +
    ggtitle(label = "Pearson Correlation")

  b <- tmp |>
    cor(use = "pairwise.complete.obs") |>
    as.data.frame() |>
    rownames_to_column(var = "Time") |>
    gather("Time2", "corr", -1) |>
    type.convert(as.is = FALSE) |>
    mutate(corr = ifelse(Time < Time2, NA, corr)) |>
    mutate(Time2 = as.factor(Time2)) |>
    ggplot(
      aes(x = Time, y = corr, group = Time2, color = Time2)
    ) +
    geom_point() +
    geom_line() +
    theme_minimal(base_size = 15) +
    color_palette(palette = "jco") +
    labs(color = "Time", y = "Pearson Correlation") +
    theme(legend.position = "top")
  ggarrange(a, b)
}
```

## Modeling

Let's fit several models with different variance-covariance structures:

```{r, eval = FALSE}
# Identity variance model.
model_0 <- asreml(
  fixed = y ~ Time + Tmt + Tmt:Time,
  residual = ~ id(Plant):idv(Time),
  data = grassUV
)

# Simple correlation model; homogeneous variance form.
model_1 <- asreml(
  fixed = y ~ Time + Tmt + Tmt:Time,
  residual = ~ id(Plant):corv(Time),
  data = grassUV
)

# Exponential (or power) model; homogeneous variance form.
model_2 <- asreml(
  fixed = y ~ Time + Tmt + Tmt:Time,
  residual = ~ id(Plant):expv(Time),
  data = grassUV
)

# Exponential (or power) model; heterogeneous variance form.
model_3 <- asreml(
  fixed = y ~ Time + Tmt + Tmt:Time,
  residual = ~ id(Plant):exph(Time),
  data = grassUV
)

# Antedependence variance model of order 1
model_4 <- asreml(
  fixed = y ~ Time + Tmt + Tmt:Time,
  residual = ~ id(Plant):ante(Time),
  data = grassUV
)

# Autoregressive model of order 1; homogeneous variance form.
model_5 <- asreml(
  fixed = y ~ Time + Tmt + Tmt:Time,
  residual = ~ id(Plant):ar1v(Time),
  data = grassUV
)

# Autoregressive model of order 1; heterogeneous variance form.
model_6 <- asreml(
  fixed = y ~ Time + Tmt + Tmt:Time,
  residual = ~ id(Plant):ar1h(Time),
  data = grassUV
)

# Unstructured variance model.
model_7 <- asreml(
  fixed = y ~ Time + Tmt + Tmt:Time,
  residual = ~ id(Plant):us(Time),
  data = grassUV
)
```


```{r, warning=FALSE, message=FALSE, echo=FALSE}
if (requireNamespace("asreml", quietly = TRUE)) {
  # Identity variance model.
  model_0 <- asreml(
    fixed = y ~ Time + Tmt + Tmt:Time,
    residual = ~ id(Plant):idv(Time),
    data = grassUV
  )

  # Simple correlation model; homogeneous variance form.
  model_1 <- asreml(
    fixed = y ~ Time + Tmt + Tmt:Time,
    residual = ~ id(Plant):corv(Time),
    data = grassUV
  )

  # Exponential (or power) model; homogeneous variance form.
  model_2 <- asreml(
    fixed = y ~ Time + Tmt + Tmt:Time,
    residual = ~ id(Plant):expv(Time),
    data = grassUV
  )

  # Exponential (or power) model; heterogeneous variance form.
  model_3 <- asreml(
    fixed = y ~ Time + Tmt + Tmt:Time,
    residual = ~ id(Plant):exph(Time),
    data = grassUV
  )

  # Antedependence variance model of order 1
  model_4 <- asreml(
    fixed = y ~ Time + Tmt + Tmt:Time,
    residual = ~ id(Plant):ante(Time),
    data = grassUV
  )

  # Autoregressive model of order 1; homogeneous variance form.
  model_5 <- asreml(
    fixed = y ~ Time + Tmt + Tmt:Time,
    residual = ~ id(Plant):ar1v(Time),
    data = grassUV
  )

  # Autoregressive model of order 1; heterogeneous variance form.
  model_6 <- asreml(
    fixed = y ~ Time + Tmt + Tmt:Time,
    residual = ~ id(Plant):ar1h(Time),
    data = grassUV
  )

  # Unstructured variance model.
  model_7 <- asreml(
    fixed = y ~ Time + Tmt + Tmt:Time,
    residual = ~ id(Plant):us(Time),
    data = grassUV
  )
}
```

## Model Comparison

We can use the Akaike Information Criterion (AIC)(Akaike, 1974) or the Bayesian Information Criterion (BIC)(Stone, 1979) for comparing the fitted models. A lower AIC or BIC value indicates a better fit.

```{r, eval=FALSE}
models <- list(
  "idv" = model_0,
  "corv" = model_1,
  "expv" = model_2,
  "exph" = model_3,
  "ante" = model_4,
  "ar1v" = model_5,
  "ar1h" = model_6,
  "us" = model_7
)

summary_models <- data.frame(
  model = names(models),
  aic = unlist(lapply(models, \(x) summary(x)$aic)),
  bic = unlist(lapply(models, \(x) summary(x)$bic)),
  loglik = unlist(lapply(models, \(x) summary(x)$loglik)),
  nedf = unlist(lapply(models, \(x) summary(x)$nedf)),
  param = unlist(lapply(models, \(x) attr(summary(x)$aic, "param"))),
  row.names = NULL
)

summary_models |> print()

summary_models |>
  ggplot(
    aes(x = reorder(model, -bic), y = bic, group = 1)
  ) +
  geom_point(size = 2) +
  geom_text(aes(x = model, y = bic + 5, label = param), size = 5) +
  geom_line() +
  theme_minimal(base_size = 15) +
  labs(x = NULL, y = "BIC")
```


```{r, warning=FALSE, message=FALSE, fig.width = 9, echo=FALSE}
if (requireNamespace("asreml", quietly = TRUE)) {
  models <- list(
    "idv" = model_0,
    "corv" = model_1,
    "expv" = model_2,
    "exph" = model_3,
    "ante" = model_4,
    "ar1v" = model_5,
    "ar1h" = model_6,
    "us" = model_7
  )

  summary_models <- data.frame(
    model = names(models),
    aic = unlist(lapply(models, \(x) summary(x)$aic)),
    bic = unlist(lapply(models, \(x) summary(x)$bic)),
    loglik = unlist(lapply(models, \(x) summary(x)$loglik)),
    nedf = unlist(lapply(models, \(x) summary(x)$nedf)),
    param = unlist(lapply(models, \(x) attr(summary(x)$aic, "param"))),
    row.names = NULL
  )

  summary_models |> print()

  summary_models |>
    ggplot(
      aes(x = reorder(model, -bic), y = bic, group = 1)
    ) +
    geom_point(size = 2) +
    geom_text(aes(x = model, y = bic + 5, label = param), size = 5) +
    geom_line() +
    theme_minimal(base_size = 15) +
    labs(x = NULL, y = "BIC")
}
```

In this specific scenario, the antedependence model emerges as the optimal choice, as indicated by the Bayesian Information Criteria (BIC). The 1-factor antedependence structure elegantly models the variance-covariance matrix $\Sigma^{\omega \times\omega}$ with the following decomposition:

$$ \Sigma ^{-1} = UDU' $$
where $U^{\omega \times\omega}$ is a unit upper triangular matrix and $D = diag(d_1, ..., d_{\omega})$ is a diagonal matrix. 

\begin{array}{rcl}
U_{ii} & = & 1 \\
U_{ij} & = & u_{ij}, \;\; 1 \le j-i\le order \\
U_{ij} & = & 0, \;\; i>j
\end{array}

and the order in our case is 1.

The `extract_rcov()` retrieves these matrices for a closer inspection of the results.

## Wald Test 

The table below shows the summary of Wald statistics for fixed effects for the models fitted.

```{r, echo=FALSE}
if (requireNamespace("asreml", quietly = TRUE)) {
  library(gt)
  models |>
    lapply(\(x) wald(x, denDF = "numeric", ssType = "conditional")$Wald) |>
    lapply(\(x) rownames_to_column(x, "Factor")) |>
    rbindlist(idcol = "Model") |>
    filter(Factor %in% c("Time", "Tmt", "Tmt:Time")) |>
    select(Model, Factor, F.con, Pr) |>
    pivot_wider(names_from = Factor, values_from = c(F.con, Pr)) |>
    mutate_if(is.numeric, round, 3) |>
    gt() |>
    tab_spanner(
      label = "Time",
      columns = c(F.con_Time, Pr_Time)
    ) |>
    tab_spanner(
      label = "Tmt",
      columns = c(F.con_Tmt, Pr_Tmt)
    ) |>
    tab_spanner(
      label = "Tmt:Time",
      columns = c(`F.con_Tmt:Time`, `Pr_Tmt:Time`)
    ) |>
    data_color(
      columns = c(`Pr_Tmt:Time`),
      method = "numeric",
      palette = "ggsci::red_material",
      domain = c(0, 0.08),
      reverse = FALSE
    ) |>
    cols_align(
      align = "center",
      columns = everything()
    ) |>
    cols_label(
      F.con_Tmt = "F.value",
      Pr_Tmt = "P.value",
      `F.con_Tmt:Time` = "F.value",
      `Pr_Tmt:Time` = "P.value",
      F.con_Time = "F.value",
      Pr_Time = "P.value"
    )
}
```
## Variance Components

At first glance, the table below looks challenging to interpret; however, the function translates the summary output into tangible forms—both the actual variance-covariance matrix and the correlation matrix. 

```{r, eval=FALSE}
summary(model_4)$varcomp
```

```{r, echo=FALSE}
if (requireNamespace("asreml", quietly = TRUE)) {
  summary(model_4)$varcomp |> print()
}
```

## Extracting Variance Covariance Matrix

Finally, to extract the variance-covariance matrix, let's take the best model according to the BIC and run the code:

```{r, eval=FALSE}
mat <- extract_rcov(model_4, time = "Time", plot = "Plant")
print(mat)
```

```{r, warning=FALSE, message=FALSE, fig.width = 9, echo=FALSE}
if (requireNamespace("asreml", quietly = TRUE)) {
  mat <- extract_rcov(model_4, time = "Time", plot = "Plant")
  print(mat)
}
```

```{r, eval=FALSE}
# Plot Correlation  Matrix
p1 <- covcor_heat(matrix = mat$corr, legend = "none", size = 4.5) +
  ggtitle(label = "Correlation Matrix (ante)")
p1

# Plot Variance-Covariance Matrix
p2 <- covcor_heat(
  matrix = mat$vcov,
  corr = FALSE,
  legend = "none",
  size = 4.5,
  digits = 1
) +
  ggtitle(label = "Covariance Matrix (ante)")
p2
ggarrange(p1, p2)
```


```{r, warning=FALSE, message=FALSE, fig.width = 9, echo=FALSE}
if (requireNamespace("asreml", quietly = TRUE)) {
  # Plot Correlation  Matrix
  p1 <- covcor_heat(matrix = mat$corr, legend = "none", size = 4.5) +
    ggtitle(label = "Correlation Matrix (ante)")
  p1

  # Plot Variance-Covariance Matrix
  p2 <- covcor_heat(
    matrix = mat$vcov,
    corr = FALSE,
    legend = "none",
    size = 4.5,
    digits = 1
  ) +
    ggtitle(label = "Covariance Matrix (ante)")
  p2
  ggarrange(p1, p2)
}
```



## Matrix Comparison

The plot below compares the raw correlation matrix with the one derived
post-application of the antedependence model.

```{r, eval=FALSE}
ggarrange(a, p1)
```

```{r, warning=FALSE, message=FALSE, fig.width = 9, echo=FALSE}
if (requireNamespace("asreml", quietly = TRUE)) {
  ggarrange(a, p1)
}
```

## Final Results

```{r, eval = FALSE}
pvals <- predict(model_4, classify = "Tmt:Time")$pvals
grassUV |>
  ggplot(
    aes(x = Time, y = y, group = Tmt, color = Tmt, shape = Tmt)
  ) +
  geom_point(alpha = 0.4, size = 3) +
  geom_line(data = pvals, mapping = aes(y = predicted.value)) +
  theme_minimal(base_size = 15) +
  color_palette(palette = "jco")
```


```{r, warning=FALSE, message=FALSE, fig.width = 9, echo = FALSE}
if (requireNamespace("asreml", quietly = TRUE)) {
  pvals <- predict(model_4, classify = "Tmt:Time")$pvals
  grassUV |>
    ggplot(
      aes(x = Time, y = y, group = Tmt, color = Tmt, shape = Tmt)
    ) +
    geom_point(alpha = 0.4, size = 3) +
    geom_line(data = pvals, mapping = aes(y = predicted.value)) +
    theme_minimal(base_size = 15) +
    color_palette(palette = "jco")
}
```

## References

[ASReml-R Reference Manual](https://asreml.kb.vsni.co.uk/wp-content/uploads/sites/3/ASReml-R-Reference-Manual-4.2.pdf)
